Principles of Modelling and Simulation in Epidemiology
===========
Laboratory exercise 2
-----------

### Suggested solutions by
Andreas Karlsson, November 2013

More information regarding Gillespie Stochastic Simulation Algorithm at **http://www.jstatsoft.org/v25/i12/paper**


```{r message=FALSE}
require(deSolve) #For the deterministic solutions (also for initial value and environmental stochasticity)
require(GillespieSSA) #Gillespie Stochastic Simulation Algorithm with "Explicit tau-leap"
require(ggplot2) #Extravagant plotting tool (not necessary to answer the questions)
require(plyr) #Handy data manipulation tool (not necessary to answer the questions)
```
### 2.1 RNGs
#### Uniform Distribution
```{r Uniform, results='hide', fig.width=7, fig.height=4}
layout(matrix(c(1,2), 1, 2, byrow = TRUE))
unif <- runif(n = 10000, min = 5, max = 15)
plot(unif, ylim = c(0,20), pch='.') + abline(h = mean(unif), col = 2, lwd = 3)
hist(unif)
```
#### Exponential Distribution
```{r Exponential, results='hide', fig.width=7, fig.height=4} 
layout(matrix(c(1,2), 1, 2, byrow = TRUE))
expo <- rexp(n = 10000, rate = 4)
plot(expo, pch='.') + abline(h = mean(expo), col = 2, lwd = 3)
hist(expo)
```
---
#### Normal Distribution
```{r Normal, results='hide', fig.width=7, fig.height=4}
layout(matrix(c(1,2), 1, 2, byrow = TRUE))
norm <- rnorm(n = 10000, mean = 2.5, sd = 10)
plot(norm, pch='.') + abline(h = mean(norm), col = 2, lwd = 3)
hist(norm)
```
#### Poisson Distribution
```{r Poisson, results='hide', fig.width=7, fig.height=4}
layout(matrix(c(1,2), 1, 2, byrow = TRUE))
po <- rpois(n = 10000,lambda = 3.5)
plot(po) + abline(h = mean(po), col = 2, lwd = 3)
hist(po)
```
---
### 2.2 Construction of other RNGs
#### Binomial Distribution direct method
```{r Binomial_direct, results='hide', fig.width=7, fig.height=4}
layout(matrix(c(1,2), 1, 2, byrow = TRUE))
bin <- rbinom(n = 100, size = 20, prob =0.3)
plot(bin) + abline(h = mean(bin), col = 2, lwd = 3)
hist(bin,breaks=0:20)
```
#### Binomial Distribution from Uniform Distribution
```{r Binomial_uniform, results='hide', fig.width=7, fig.height=4}
layout(matrix(c(1,2), 1, 2, byrow = TRUE))
all_ber <- NULL
for (i in 1:20){
  unif <- runif(n = 100, min = 0, max = 1)
  ber <- ifelse(unif<0.3, 1, 0)
  all_ber <- rbind(all_ber,ber)
}
bin <- colSums(all_ber)
plot(bin) + abline(h = mean(bin), col = 2, lwd = 5)
hist(bin,breaks=0:20)
```
---
### 2.3 RNG and time-step
```{r RNG_dt, results='hide', fig.width=7, fig.height=4}
rm(list=ls(all=TRUE))
layout(matrix(c(1:6), 2, 3, byrow = TRUE))
for (dt in c(1, 0.1)) {
  derivs <- function(times, state, parameters) {
    with(as.list(c(state, parameters)), {
      dU <- unif[times / dt] - c1 * U
      dUout <-  c1 * U
      return(list(c(dU,dUout)))
    })
  }
  init <- c(U = 0, Uout = 0)
  times <- seq(1, 100, by = dt)
  unif <- runif(n = length(times), min = 0, max = 10)
  parameters <-  c(unif, c1 = 0.1, dt)
  out <- as.data.frame(ode(y = init, times = times,
                           func = derivs, parms = parameters))
  
  ## Plot results
  plot(times, unif, type = 'l', xlab = 'Time', ylab = 'AU / t',
       main = paste('In_Unif, ',sprintf("dt=%.1f", dt)), lty = 1, bty = 'l', col = 2)
  plot(times, out$U, type = 'l', xlab = 'Time', ylab = 'AU',
       main = paste('U, ',sprintf("dt=%.1f", dt)), lty = 1, bty = 'l', col = 2)
  plot(times, out$Uout/times, type = 'l', xlab = 'Time', ylab = 'AU / t',
       main = paste('Out_U, ',sprintf("dt=%.1f", dt)), lty = 1, bty = 'l', col = 2)
}
```
### 2.4 Seeds - making a stochastic simulation reproducible
```{r Seeds, results='hide', fig.width=7, fig.height=4}
rm(list=ls(all=TRUE))
layout(matrix(c(1:2), 1, 2, byrow = TRUE))
set.seed(1234)
un1 <- runif(n = 100, min = 0, max = 10)
set.seed(1234)
un2 <- runif(n = 100, min = 0, max = 10)
plot(1:100, un1, type = 'l', xlab = 'Time', ylab = 'AU',
     main = 'Uniform 1', lty = 1, bty = 'l', col = 2)
plot(1:100, un2, type = 'l', xlab = 'Time', ylab = 'AU',
     main = 'Uniform 2', lty = 1, bty = 'l', col = 2)
```
---
### 3.1 Demographic Stochasticity
```{r Demog, results='hide', fig.width=7, fig.height=4}
rm(list=ls(all=TRUE)) 
layout(matrix(1:2, 1, 2))
init <- c(X = 1)
parms <- c(c1 = 1, c2 = 0.01)
deriv <- function(times, state, parameters) {
  with(as.list(c(state, parameters)), {
    dX <- c1 * X - c2 * X * X
    return(list(c(dX)))
  })
}
times <- seq(0, 20, by = 0.01)
out <- as.data.frame(ode(y = init, times = times, func = deriv, parms = parms))
out$time <- NULL
matplot(times, out , type = 'l', xlab = 'Time', ylab = 'AU', ylim = c(0, 120),
        main = 'Deterministic Logistic Model', lty = 1, lwd =1, col = 2)

#Gillespie with "Explicit tau-leap" => user-defined step size 
a <- c("c1 * X", "c2 * X * X")
nu <- matrix(c(+1,-1),ncol=2)
out <- ssa(init, a , nu, parms,
           tf=20,
           tau = 0.01,
           method="ETL",
           maxWallTime=5,
           simName='Stochastic Logistic Model')
ssa.plot(out,show.legend=F)
```
---
### 3.2 Environmental Stochasticity
```{r Enviro, results='hide', fig.width=7, fig.height=4}
rm(list=ls(all=TRUE)) 
layout(matrix(1:2, 1, 2))
for (f in c(5,0.1)){
  deriv <- function(Time, state, parameters) {
    with(as.list(c(state, parameters)), {
      set.seed(1000 + ceiling(Time/f)) #Shifting seed at an interval
      c1 <- runif(1, 0.5 * c1, 1.5 * c1)
      set.seed(2000 + ceiling(Time/f))
      c2 <- rnorm(1, c2, 0.2 * c2)
      dX <- c1 * X - c2 * X * X
      return(list(c(dX)))
    })
  }
  
  init <- c(X = 1)
  times <- seq(0, 20, by =0.01)
  parameters <- c(c1 = 1, c2 = 0.01, f)
  out <- as.data.frame(ode(y = init, times = times,
                           func = deriv, parms = parameters))
  out$time <- NULL
  matplot(times, out , type = 'l', xlab = 'Time', ylab = 'AU', ylim = c(0, 120),
          main = paste('Environmental Stochasticity\n',
                       sprintf("Interval=%.1f", f)), lty = 1, lwd =1, col = 2)
}
```
### 3.3 Initial Value Stochasticity
```{r Initial, results='hide', fig.width=7, fig.height=4}
rm(list=ls(all=TRUE)) 
layout(matrix(1:2, 1, 2))
for (seed_n in c(1,1000)){
  set.seed(seed_n)
  deriv <- function(Time, state, parameters) {
    with(as.list(c(state, parameters)), {
      dX <- c1 * X - c2 * X * X
      return(list(c(dX)))
    })
  }  
  init <- c(X = rbinom(1, 5, 0.2))
  times <- seq(0, 20, by =0.01)
  parameters <- c(c1 = 1, c2 = 0.01)
  out <- as.data.frame(ode(y = init, times = times,
                           func = deriv, parms = parameters))
  out$time <- NULL
  matplot(times, out , type = 'l', xlab = 'Time', ylab = 'AU', ylim = c(0, 120),
          main = 'Initial Value Stochasticity', lty = 1, lwd =1, col = 2)
}
```
---
### 4.1 Output within replication
```{r Output_within, results='hide', fig.width=7, fig.height=4}
rm(list=ls(all=TRUE))
layout(matrix(1:2, 1, 2))
parms <- c(c1 = 1, c2 = 0.01)
x0 <- c(X=100)
a <- c("c1 * X", "c2 * X * X")
nu <- matrix(c(+1,-1),ncol=2)
sim_out <- data.frame(dt = character(), X = numeric())
for (dt in c(0.1,0.01)){
  #"Explicit tau-leap" => user-defined step size
  out <- ssa(x0, a , nu, parms,
             tf=100,
             tau = dt,
             method="ETL",
             maxWallTime=5,
             simName='Stochastic Logistic Model',
             verbose = F)
  ## Plot results
  ssa.plot(out,show.legend=F) + abline(h = mean(out$data[,2]), col = 1) +
    abline(h = mean(out$data[,2]) + sd(out$data[,2]), col = 1, lty = 3) +
    abline(h = mean(out$data[,2]) - sd(out$data[,2]), col = 1, lty = 3)
  
  sim_out <- rbind(sim_out, data.frame(dt = rep(dt,nrow(out$data) ), X = out$data[,2]))
}
```
```{r Output_within_txt}
dt_summay <- ddply(sim_out, .(dt), summarise, Lowest_X = min(X), Highest_X = max(X), StdDev_X = round(sd(X),0))
print(dt_summay)
```
---
### 4.2 Output over many replications
```{r Output_many, results='hide', fig.width=11, fig.height=7}
rm(list=ls(all=TRUE))
sim_out <- data.frame(Repetition = numeric(), Sample = numeric(), Time = double(), X = double())
parms <- c(c1 = 1, c2 = 0.01)
x0 <- c(X=100)
a <- c("c1 * X", "c2 * X * X")
nu <- matrix(c(+1,-1),ncol=2)
for (i in c(1:1000)){
  #"Explicit tau-leap" => user-defined step size
  out <- ssa(x0, a , nu, parms,
             tf=100,
             tau = 0.1,
             method="ETL",
             maxWallTime=5)
  colnames(out$data)[1] <- 'Time'
  Repetition <- rep(i,nrow(out$data))
  Sample <- 1:nrow(out$data)
  sim_out <- rbind(sim_out,cbind(Repetition,Sample,out$data))
}
within_summary <- ddply(sim_out, .(Repetition), summarise, X_end = tail(X, n=1), X_max = max(X))

## Plot results
ggplot(data = sim_out, aes(x=Time, y=X, colour=Repetition)) + geom_line(aes(group = Repetition))
```
---
```{r Output_many_2, results='hide', fig.width=11, fig.height=7}
median_quantiles <- quantile(within_summary$X_end, c(.025, 0.50, .975))
within_summary$Percentile <- 'middle'
within_summary$Percentile[within_summary$X_end < median_quantiles['2.5%']] <- '<2.5%'
within_summary$Percentile[within_summary$X_end > median_quantiles['97.5%']] <- '>97.5%'

mean_sd <- data.frame(mean=mean(within_summary$X_end), sd=sd(within_summary$X_end))
ggplot(data = within_summary, aes(x=Repetition, y=X_end, colour = Percentile))  +
  geom_point(aes(group = Percentile)) + ggtitle('X(100), Mean & 95% CI') +
  geom_hline(data=mean_sd,aes(yintercept=mean,3),linetype=1, colour="#990000") +
  geom_hline(data=mean_sd,aes(yintercept=mean+1.9604*sd,3),linetype=2, colour="#990000") +
  geom_hline(data=mean_sd,aes(yintercept=mean-1.9604*sd,3),linetype=2, colour="#990000") +
  ylab("X(100)")
```
#### Mean and CI for X[100]:
```{r Output_many_txt1}
print(paste('Mean:', round(mean_sd$mean),'  95% CI[', round(mean_sd$mean - mean_sd$sd * 1.9604),
            ',' , round(mean_sd$mean + mean_sd$sd * 1.9604),']'))
```
#### Corresponding percentiles:
```{r Output_many_txt2}
print(median_quantiles)
```
#### The maximum of the end values:
```{r Output_many_txt3}
print(paste('Max[X(100)]:', max(within_summary$X_end)))
```
#### The maximum of the highest value within each repetition:
```{r Output_many_txt4}
print(paste('Max[Highest_X]:', max(within_summary$X_max)))
```
---

### 5 Comparing Logistic & SI models
```{r Logistic_SI, warning=FALSE, results='hide', fig.width=11, fig.height=7}
rm(list=ls(all=TRUE))
sim_out <- data.frame(Repetition = numeric(), Sample = numeric(),
                      Time = numeric(), X = numeric(), Model = character())

## Defining constants
parms <- c(c1 = 1, c2 = 0.1, r=0.1)
init <- c(X = 1, S=99, I = 1)

## Deterministic Logistic & SI model
deriv <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    dX <- c1 * X - c2 * X * X
    dS <-   - r * I * S
    dI <-     r * I * S
    return(list(c(dX, dS, dI)))
  })
}
times <- seq(0, 20, by = 0.01)
out <- as.data.frame(ode(y = init, times = times,
                         func = deriv, parms = parms))
Repetition <- rep(1,nrow(out)) 
Sample <- 1:nrow(out)
sim_out <- rbind(sim_out, data.frame(Repetition, Sample, Time = times, X = out$X, Model = rep('Deterministic Logistic Model',nrow(out))))   
sim_out <- rbind(sim_out, data.frame(Repetition, Sample, Time = times, X = out$I,  Model = rep('Deterministic SI Model',nrow(out))))

## Stochastic Logistic & SI model
for (i in 1:1000){
  a <- c("c1 * X", "c2 * X * X", "r*S*I")
  nu <- matrix(c(+1, -1, 0, 0, 0, -1, 0, 0, +1), nrow=3, byrow=T)
  #"Explicit tau-leap" => user-defined step size
  out <- ssa(init, a , nu, parms,
             tf=20,
             tau = 0.01,
             method="ETL",
             maxWallTime=5,
             ignoreNegativeState = T) #Gives warning instead of error for negative number of susceptibles.
  Repetition <- rep(i,nrow(out$data))
  Sample <- 1:nrow(out$data)
  sim_out <- rbind(sim_out, data.frame(Repetition, Sample, Time = out$data[,1], X = out$data[,2], 
                                       Model = rep('Stochastic Logistic Model',nrow(out$data))))
  sim_out <- rbind(sim_out, data.frame(Repetition, Sample, Time = out$data[,1], X = out$data[,4],
                                       Model = rep('Stochastic SI Model',nrow(out$data))))
}
within_summary <- ddply(sim_out, .(Model, Repetition), summarise, X_end = tail(X, n=1))

## Plot results
ggplot(sim_out, aes(x=Time, y=X, colour=Repetition)) +
  geom_line(aes(group = Repetition)) +
  facet_wrap( ~ Model, ncol=2)
```
---
```{r Logistic_SI_2, warning=FALSE, results='hide', fig.width=11, fig.height=7}
mean_sd <- ddply(within_summary,.(Model), summarise,
                 mean=mean(X_end, na.rm = TRUE), sd=sd(X_end, na.rm = TRUE))
ggplot(data = within_summary, aes(x=Repetition, y=X_end))  + geom_point() + ylab("X(end)") +
  geom_hline(data=mean_sd,aes(yintercept=mean,3),linetype=1, colour="#990000") +
  geom_hline(data=mean_sd,aes(yintercept=mean+1.9604*sd,3),linetype=2, colour="#990000") +
  geom_hline(data=mean_sd,aes(yintercept=mean-1.9604*sd,3),linetype=2, colour="#990000") +
  facet_wrap( ~ Model, ncol=2) +
  ggtitle('X(end), Mean & 95% CI')
```
#### End sample summary:
```{r Logistic_SI_txt1}
End_summary <-ddply(within_summary, .(Model), summarise, Av_end_X = mean(X_end),
                    CI_low =mean(X_end) - 1.9604 * sd(X_end), CI_high =mean(X_end) + 1.9604 * sd(X_end) )
print(End_summary)
```
#### Percent Extinction summary:
```{r Logistic_SI_txt2}
PercentExtinct <-ddply(within_summary, .(Model), summarise,
                           PercentExtinct = length(X_end[X_end<0.5]) / length(X_end) * 100 )
print(PercentExtinct)
```
#### End sample summary (discriminating extinct):
```{r Logistic_SI_txt3}
No_ext_End_summary <- ddply(within_summary[within_summary$X_end>0.5,], .(Model), summarise,
                            Av_end_X = mean(X_end), CI_low =mean(X_end) - 1.9604 * sd(X_end),
                            CI_high =mean(X_end) + 1.9604 * sd(X_end) )
print(No_ext_End_summary)
```